Model-Based and Model-Free Reinforcement Learning in Accelerator Operation with Application to FERMI FEL

### 2020-12-07 main.tex
- Niky last fermi test

	\title{Online Model-Based and Model-Free Reinforcement Learning in Accelerator Operation with Application to FERMI FEL}

	\author{Simon Hirlaender}
	%\thanks{These two authors contributed equally}
	% \affiliation{University of Malta}
	\affiliation{Univertsity of Salzburg, Salzburg, Austria}
	\author{Niky Bruchon}%, Gianfranco Fenu, Felice Andrea Pellegrino, Erica Salvato}
	\affiliation{University of Trieste, Trieste, Italy}
	% \author{Gianluca Valentino}
	% \affiliation{University of Malta, Msida, MT}
	% \author{Marco Lonza, Giulio Gaio}
	% \affiliation{Elettra Sincrotrone Trieste, Trieste, IT}
	% \author{Verena Kain}
	% %\thanks{These two authors contributed equally}
	% \email{verena.kain@cern.ch}

@@

	%\tableofcontents
	\section{Introduction and Motivation}
	In particle accelerator operation, one main goal is to provide stable and reproducible performance. Achieving this demands to consider several control problems simultaneously \cite{Kain2020}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015,Albright2019} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance.\\
	 Recently, the community has begun to explore a novel approach to solve control problems. Reinforcement learning (RL) \cite{Bruchon2020,Bruchon2019,Kain2020,Pang2020} unveils several advantages over optimisation methods:
	\begin{itemize}
		\item It covers a larger class of problems, as RL optimises a sequence of decisions.
		\item It memorises the problem and does not always begin from zero as a DFO.
		\item Already existing data can be used.
		\item Adaption to non-stationary systems is possible.
		\item The underlying structure of the problem might be deduced.
	\end{itemize}
	One demanding aspect using RL is the number of iterations needed to train a controller - the sample-efficiency, since RL methods are known to be data-hungry \cite{Sutton2018}, and a second critical aspect is the robustness of the training in real world problems itself.\\
	In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste.\\
	 Two different algorithm classes (\emph{NAF} and \emph{AE-DYNA}) were applied successfully to the FERMI FEL problem and reveal different advantages. The critical aspects are addressed, and tailored solutions, generally applicable to problems showing a sample-complexity as in accelerator control problems are presented.
	The paper is organised as follows:
	\begin{itemize}
		\item Description of the problem set-up at FERMI
		\item Overview of RL
		\item Design decisions of the implementations used in these studies and theoretical concerns
		\item Results of the experiments 
		\item Summary and outlook
	\end{itemize}
	% The methods proposed are the iterative Linear Quadratic Regulator (iLQR) - an extension to non-linear systems of the fundamental control theory Linear Quadratic Regulator (LQR) problem -  and the Normalised Advantage Function (NAF) - a continuous variant of the Q-learning algorithm from reinforcement learning field. An exhaustive presentation of both techniques is provided in the following sections.

@@

 To provide the possibility for other laboratories to use the stated methods, the code was released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper} and \cite{Hirlaender2020}.
 \vfill
 \newpage
 \vfill

@@

%	\section{Theoretical aspects on the \emph{AE-DYNA} on FERMI FEL}\label{appendix:Theoretical aspects on the 	AE-DYNA on FERMI FEL}
%	In thist section we want to discuss some theoretical aspects concerning the \emph{AE-DYNA} approach. A simulation was used to obtain the presented analysis. \\
%	One issue of MBRL methods is asymptotic performance. Although good results were obtained using the \emph{AE-DYNA}, there is a difference between the MFRL and MBRL. We call it the \emph{reality gap}. There are ideas to attack the problem, e.g. by learning a meta-policy \cite{Clavera2018} followed by a fine-tuning on the residual physics, which minimises the \emph{reality gap} of training on a simulator and closing than with a small number of training iterations on the real system the gap \cite{Zeng2019}.
%	\begin{figure}[!h]
%		\centering
%		\includegraphics*[width=0.5\textwidth]{Figures/Learning_evolution}
%		\caption{The training .}
%		\label{fig:Learning_evolution}
%	\end{figure}
\section{Still open}
	\begin{itemize}
	\item Reference and code clean up.
	\item Conclusion and outlook\dots
	\item Clean discussion of the results. What is superior and why? A table with numbers \dots
	\item Add the noise policy of NAF
	\item Check the plots, which might be already okay!
%	\item Noise benchmarks with pendulum - done
%	\item AE details from papers
%	\item Benchmark NAF2 with pendulum - done
%	\item Compare NAF2 and AE-DYNA
%	\item The first time model based on accelerator problem compared to a novel state of the art MFRL: NAF2
%	\item Tailored to accelerators - short horizons are standard?
%	\item Long waiting time in MBRL	
%	\item Vary the model size
%	\item Noise sensitivity as mentioned in Benchmark paper
%	\item Improvements of the algorithms:
%	\begin{itemize}
%		\item Adding the noise feature
%		\item Jumping feature using \emph{SAC}
%		\item Noise in the policy
%		\item discuss different approaches to, e.g. dynamic waiting time
%		\item usually prior not discussed in papers
%	\end{itemize}
\end{itemize}

>>>>

	\title{Model-Based and Model-Free Reinforcement Learning in Accelerator Operation with Application to FERMI FEL}

	\author{Simon Hirlaender}
	%\thanks{These two authors contributed equally}
	% \affiliation{University of Malta}
	\affiliation{University of Salzburg, Salzburg, Austria}
	\author{Niky Bruchon}%, Gianfranco Fenu, Felice Andrea Pellegrino, Erica Salvato}
	\affiliation{University of Trieste, Trieste, Italy}
	% \author{Gianluca Valentino}
	% \affiliation{University of Malta, Msida, MT}
	% \author{Marco Lonza, Giulio Gaio}
	% \affiliation{Elettra Sincrotrone Trieste, Trieste, IT}
	% \author{Verena Kain}
	% %\thanks{These two authors contributed equally}
	% \email{verena.kain@cern.ch}

@@

	%\tableofcontents
	\section{Introduction and Motivation}
	In particle accelerator operation, one main goal is to provide stable and reproducible performance. Achieving this demands to consider several control problems simultaneously \cite{Kain2020}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015,Albright2019} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance.\\
	 Recently, the community has begun to explore a novel approach to solve control problems. Reinforcement learning (RL) \cite{Bruchon2020,Bruchon2019,Kain2020,Pang2020} unveils several advantages over optimisation methods:
	\begin{itemize}
		\item It covers a larger class of problems, as RL optimises a sequence of decisions.
		\item It memorises the problem and does not always begin from zero as a DFO.
		\item Already existing data can be used.
		\item Adaption to non-stationary systems is possible.
		\item The underlying structure of the problem might be deduced.
	\end{itemize}
	One demanding aspect using RL is the number of iterations needed to train a controller - the sample-efficiency, since RL methods are known to be data-hungry \cite{Sutton2018,DulacArnold2019}, and a second critical aspect is the robustness of the training in real world problems itself.\\
	In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste.\\
	 Two different algorithm classes (\emph{NAF} and \emph{AE-DYNA}) were applied successfully to the FERMI FEL problem and reveal different advantages. The critical aspects are addressed, and tailored solutions, generally applicable to problems showing a sample-complexity as in accelerator control problems are presented.
	The paper is organised as follows:
	\begin{itemize}
		\item Description of the problem set-up at FERMI
		\item Overview of RL
		\item Design decisions of the implementations used in these studies and theoretical concerns
		\item Results of the experiments 
		\item Summary and outlook
	\end{itemize}
	% The methods proposed are the iterative Linear Quadratic Regulator (iLQR) - an extension to non-linear systems of the fundamental control theory Linear Quadratic Regulator (LQR) problem -  and the Normalised Advantage Function (NAF) - a continuous variant of the Q-learning algorithm from reinforcement learning field. An exhaustive presentation of both techniques is provided in the following sections.

@@

 To provide the possibility for other laboratories to use the stated methods, the code was released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper} and \cite{Hirlaender2020b}.
 \vfill
 \newpage
 \vfill

@@

%	\section{Theoretical aspects on the \emph{AE-DYNA} on FERMI FEL}\label{appendix:Theoretical aspects on the 	AE-DYNA on FERMI FEL}
%	In thist section we want to discuss some theoretical aspects concerning the \emph{AE-DYNA} approach. A simulation was used to obtain the presented analysis. \\
%	One issue of MBRL methods is asymptotic performance. Although good results were obtained using the \emph{AE-DYNA}, there is a difference between the MFRL and MBRL. We call it the \emph{reality gap}. There are ideas to attack the problem, e.g. by learning a meta-policy \cite{Clavera2018} followed by a fine-tuning on the residual physics, which minimises the \emph{reality gap} of training on a simulator and closing than with a small number of training iterations on the real system the gap \cite{Zeng2019}.
%	\begin{figure}[!h]
%		\centering
%		\includegraphics*[width=0.5\textwidth]{Figures/Learning_evolution}
%		\caption{The training .}
%		\label{fig:Learning_evolution}
%	\end{figure}
\section{Still open}
	\begin{itemize}
	\item Reference and code clean up.
	\item Conclusion and outlook\dots
	\item Clean discussion of the results. What is superior and why? A table with numbers \dots
	\item Add the noise policy of NAF
	\item Check the plots, which might be already okay!
	\item Acknowledgements
%	\item Noise benchmarks with pendulum - done
%	\item AE details from papers
%	\item Benchmark NAF2 with pendulum - done
%	\item Compare NAF2 and AE-DYNA
%	\item The first time model based on accelerator problem compared to a novel state of the art MFRL: NAF2
%	\item Tailored to accelerators - short horizons are standard?
%	\item Long waiting time in MBRL	
%	\item Vary the model size
%	\item Noise sensitivity as mentioned in Benchmark paper
%	\item Improvements of the algorithms:
%	\begin{itemize}
%		\item Adding the noise feature
%		\item Jumping feature using \emph{SAC}
%		\item Noise in the policy
%		\item discuss different approaches to, e.g. dynamic waiting time
%		\item usually prior not discussed in papers
%	\end{itemize}
\end{itemize}
